Evolution of Processes in Delivering Reliable Software within the European Scientific Arena

### 2019-02-11 paper-jgc.tex
added file for the new template

\documentclass[journal]{IEEEtran}

\usepackage{graphicx}
\usepackage{array}
\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage{url}
\usepackage{cite}

\begin{document}
\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena}

\author{E. Ronchieri,
        P. Orviz Fernandez,
        M. David,
        D. C. Duma,
        J. Gomes,
        D. Salomoni
\thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}
\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain.}
\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal.}%
}

@@

\begin{abstract}

From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering processes used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely constituted a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidence of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.

\end{abstract}

\begin{IEEEkeywords}
Software Reliability, Quality Assurance, Software Metrics, Software Testing
Techniques, DevOps
\end{IEEEkeywords}

\IEEEpeerreviewmaketitle
\section{Introduction}

\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
supporting unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator.

A significant number of innovation projects and technology transfer
activities -- later discussed in Section~\ref{sec:ev} --
were funded by the 5th, 6th and 7th Framework Programmes of the European Commission (EC)
and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
software solutions that provide a seamless and transparent access by researchers to
distributed computing and data resources.

Major developments carried throughout the years were focused on providing new
functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
% Comment
%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

% Comments
%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
In an effort to improve the research community's experience, it became apparent the need for a better
balance between the addition of new features and the reliability of the released software. Consequently,
incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the
software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

@@

\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{List of EC--projects}
\label{tab:eup}
\centering
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E--sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI--Inspire &
Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

\begin{table*}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &$\surd$&       &$\surd$&$\surd$&$\surd$&$\surd$\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
\end{table*}

@@

Project monitoring and reporting mechanisms allowed to assess the risks, efforts and documentation
produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
1) packaging, 2) test and validation, and 3) style and naming convention. The specification of quality indicators
\cite{datagridindicators} facilitated the detection of deviations in the collected metric data. Figure
\ref{fig:efficiency} shows the efficiency performance indicator, measured in production environments as the
number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
there is a positive trend of efficiency over the months, with the unique exception of the LCG1 experiment.
\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/datagridEff.png}
\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
\label{fig:efficiency}
\end{figure}

@@

\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{EMI software quality evolution}
\label{tab:emi}
\centering
\begin{tabular}{lp{1.6cm}p{1.6cm}p{1.3cm}p{1.5cm}}
\hline
\hline
PQ & Release Products (RP) & N. RP Passed UMD QC & N. RP Passed SR & N. RP Failed UMD QC\\
\hline
\hline
5 & 30 & 27 & 27 & 0\\
6 & 30 & 28 & 26 & 2 \\
7 & 27 & 26& 24& 2\\
8 & 18 & 18 & 18 & 0\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth]{images/codestyle.png}
\caption{Code style standards followed by {\sl INDIGO--DataCloud}'s software products.}
\label{fig:fig_codestyle}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth]{images/fig3.png}
\caption{Adoption of Configuration Management tools throughout the project lifetime. The figure shows the trend lines
leading to the first (light cream points and line) and second release (dark blue points and line).}
\label{fig:fig_confman}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{Evolution of the total number of testing builds triggered automatically as part of the Jenkins CI implementation.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/devops.png}
\caption{Continuous Delivery workflow for Docker images.}
\label{fig:fig_CD}
\end{figure*}

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stages surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project.
\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of defects detected in development and production stages over the lifetime of the
{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
stages, thus not affecting the software released.}
\label{fig:indigobugs}
\end{figure}

@@

\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{array}
\usepackage{subfigure}
\usepackage{adjustbox}
\usepackage{listings}
\usepackage{url}
\usepackage{cite}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{}

%\titlerunning{Short form of title}        % if too long for running head

\author{M. David \and
        D. C. Duma \and
        P. Orviz Fernandez \and
        E. Ronchieri \and
        J. Gomes \and
        D. Salomoni
%
%First Author         \and
%        Second Author %etc.
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{M. David \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{david@lip.pt}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
\and
D. C. Duma \at
INFN - CNAF, Bologna, Italy\\
\email{cristina.aiftimiei@cnaf.infn.it}
\and
P. Orviz Fernandez \at
CSIC, Santander, Spain \\
\email{orviz@ifca.unican.es}
\and
J. Gomes \at
LIP, Lisbon, Portugal \\
%Tel.: +123-45-678910\\
%Fax: +123-45-678910\\
\email{jorge@lip.pt}
\and
E. Ronchieri \at
INFN - CNAF, Bologna, Italy\\
\email{elisabetta.ronchieri@cnaf.infn.it}
\and
D. Salomoni \at
INFN - CNAF, Bologna, Italy\\
\email{davide.salomoni@cnaf.infn.it}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
This work examines the evolution of the software engineering processes used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely constituted a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidence of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\keywords{Software Reliability \and Quality Assurance \and Software Metrics \and Software Testing
Techniques \and DevOps}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
During the last 15 years the provision of an European e--Infrastructure,
supporting unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator.

A significant number of innovation projects and technology transfer
activities -- later discussed in Section~\ref{sec:ev} --
were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
software solutions that provide a seamless and transparent access by researchers to
distributed computing and data resources.

Major developments carried throughout the years were focused on providing new
functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
The released software was less tested, resulting in a higher rate of
rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

In an effort to improve the research community's experience, it became apparent the need for a better
balance between the addition of new features and the reliability of the released software. Consequently,
incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered. At this point, evolving software engineering practices were considered
and gradually adopted to define and implement quality procedures, the organization of
cross--functional teams and deployment of pilot testbeds.

@@

\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{List of EC--projects}
\label{tab:eup}
\centering
\begin{tabular}{p{1.6cm}p{2cm}p{5cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E--sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI--Inspire &
Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

\begin{table*}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
\begin{tabular}{p{2.6cm}lllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition        &$\surd$&       &$\surd$&$\surd$&$\surd$&$\surd$\\
Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
\hline
\hline
\end{tabular}
\end{table*}

@@

Project monitoring and reporting mechanisms allowed to assess the risks, efforts and documentation
produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
1) packaging, 2) test and validation, and 3) style and naming convention. The specification of quality indicators
\cite{datagridindicators} facilitated the detection of deviations in the collected metric data. Figure
\ref{fig:efficiency} shows the efficiency performance indicator, measured in production environments as the
number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
there is a positive trend of efficiency over the months, with the unique exception of the LCG1 experiment.
\begin{figure}[h]
\centering
%\includegraphics[width=0.8\textwidth, height=50mm]{images/datagridEff.png}
\includegraphics[width=1\textwidth]{images/datagridEff.png}
\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
\label{fig:efficiency}
\end{figure}

@@

\begin{table}[!h]
\renewcommand{\arraystretch}{1.3}
\caption{EMI software quality evolution}
\label{tab:emi}
\centering
\begin{tabular}{lp{2.5cm}p{2.2cm}p{1.9cm}p{2.1cm}}
\hline
\hline
PQ & Release Products (RP) & N. RP Passed UMD QC & N. RP Passed SR & N. RP Failed UMD QC\\
\hline
\hline
5 & 30 & 27 & 27 & 0\\
6 & 30 & 28 & 26 & 2 \\
7 & 27 & 26& 24& 2\\
8 & 18 & 18 & 18 & 0\\
\hline
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth]{images/codestyle.png}
\caption{Code style standards followed by {\sl INDIGO--DataCloud}'s software products.}
\label{fig:fig_codestyle}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth]{images/fig3.png}
\caption{Adoption of Configuration Management tools throughout the project lifetime. The figure shows the trend lines
leading to the first (light cream points and line) and second release (dark blue points and line).}
\label{fig:fig_confman}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{Evolution of the total number of testing builds triggered automatically as part of the Jenkins CI implementation.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{images/devops.png}
\caption{Continuous Delivery workflow for Docker images.}
\label{fig:fig_CD}
\end{figure}

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are possible, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
total number of defects detected in the pre--production stages surpassed by a factor of 30
the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
project.
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
\caption{Number of defects detected in development and production stages over the lifetime of the
{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
stages, thus not affecting the software released.}
\label{fig:indigobugs}
\end{figure}

@@

%
%
%\section{Section title}
%\label{sec:1}
%Text with citations \cite{RefB} and \cite{RefJ}.
%\subsection{Subsection title}
%\label{sec:2}
%as required. Don't forget to give each section
%and subsection a unique label (see Sect.~\ref{sec:1}).
%\paragraph{Paragraph headings} Use paragraph headings as needed.
%\begin{equation}
%a^2+b^2=c^2
%\end{equation}
%
%% For one-column wide figures use
%\begin{figure}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
%  \includegraphics{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:1}       % Give a unique label
%\end{figure}
%%
%% For two-column wide figures use
%\begin{figure*}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
%  \includegraphics[width=0.75\textwidth]{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:2}       % Give a unique label
%\end{figure*}
%%
%% For tables use
%\begin{table}
%% table caption is above the table
%\caption{Please write your table caption here}
%\label{tab:1}       % Give a unique label
%% For LaTeX tables use
%\begin{tabular}{lll}
%\hline\noalign{\smallskip}
%first & second & third  \\
%\noalign{\smallskip}\hline\noalign{\smallskip}
%number & number & number \\
%number & number & number \\
%\noalign{\smallskip}\hline
%\end{tabular}
%\end{table}

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

% Non-BibTeX users please use
%\begin{thebibliography}{}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
%\bibitem{RefJ}
% Format for Journal Reference
%Author, Article title, Journal, Volume, page numbers (year)
% Format for books
%\bibitem{RefB}
%Author, Book title, page numbers. Publisher, place (year)
% etc
%\end{thebibliography}

\end{document}
% end of file template.tex
